Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-03 source/sec_problem.tex
update problem.tex

In this paper, we study the following federated learning problem:
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum\nolimits_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In $k$-th local device, there are a finite number of $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. In federated learning, each device stores its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumption on the
data distributions of each local device. The local minimum
$\vw^{*}_k= \argmin_{\vw \in R^d} F_k(\vw)$ can be drifted far from
the global minimum of \eq{\ref{eq:problem}}, which quantifies the 
heterogeneity of the problem.
In this paper, we systematically investigate the convergence results, 
when the local objectives are convex smooth, strongly convex and convex but
non-smooth.

@@

\subsection{The Federated Averaging (FedAvg) Algorithm}
We first introduce the standard Federated Average (FedAvg) algorithm.
The FedAvg updates the model in each device by local Stochastic Gradient Descent (SGD) and send the latest model to the central server every $E$
steps. The central server conducts a weighted average over the models
received from active devices and broadcasts the latest averaged model to all devices.
Formally, the updates of FedAvg at round $t$ is described as follows:
\begin{align}
\vv_{t+1}^{k} & =\vw_{t}^{k}-\alpha_{t}\vg_{t,k}\\
\mathbf{w}_{t+1}^{k} &=\left\{
\begin{array}{ll}
\mathbf{v}_{t+1}^{k} & \text { if } t+1 \notin \mathcal{I}_{E}, \\ 
\sum_{k \in \cS_{t+1}} \mathbf{v}_{t+1}^{k} & \text { if } t+1 \in \mathcal{I}_{E}
\end{array}\right.
\end{align}
where $\vw_t^k$ is the local model parameter maintained in the $k$-th device at the $t$-th iteration, $\vg_{t,k}:=\nabla F_{k}(\vw_{t}^{k},\xi_{t}^{k})$ is the stochastic gradient based on $\xi_{t}^{k}$, the data sampled from $k$-th device’s local data uniformly at random. $\mathcal{I}_{E}=\{E,2E,\dots\}$ is the set of global synchronization steps. We use $\mathcal{S}_{t+1}$ to represent the set of active devices at $t+1$.

Since federated learning usually involves an enormous amount of 
local devices, it is more realistic to consider only a subset of 
local devices is active at each communication round (system heterogeneity). In this work,
we consider both the case of \textbf{full participation} where the model is
averaging over all devices at the communication round, i.e.,
$\mathbf{w}_{t+1}^{k} = \sum_{k=1}^N p_k \mathbf{v}_{t+1}^{k}$, and
the case of \textbf{partial participation} where $|\cS_{t+1}| < N$. 
In the partial participation scenario, $\cS_{t+1}$ is obtained by two types
sampling schemes to simulate practical scenarios~\cite{li2019convergence}. 
The sampling scheme I establishes $\cS_{t+1}$ by i.i.d. sampling the devices according to
the weight $p_k$ with replacement. 
The sampling scheme II establishes $\cS_{t+1}$ by uniformly sampling all devices without
replacement. Both of the sampling schemes
guarantee that gradient updates in FedAvg are unbiased stochastic versions of
updates in FedAvg with full participation. For 
more details on the notations and setup, please refer to Section~\ref{sec:app:notations} in the appendix.

@@

\subsection{Assumptions}
\label{sec:assumptions}
We make the following standard assumptions on the objective function $F_1,\dots, F_N$. Assumption~\ref{ass:lsmooth} and Assumption~\ref{ass:stroncvx} are common assumptions satisfied by a range of popular objective functions, such as $\ell^{2}$ regularized logistic regression and cross-entropy loss functions.

@@

\textbf{Justfication for Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2}.} These standard assumptions have been made in many previous
works in federated learning\cite{yu2019parallel,li2019convergence,stich2018local}. Here we provide one more justification
for their generality: even though they require bounds for all $\mathbf{w}$,
they are not much stronger than requiring bounds at the global optimal
parameter $\mathbf{w}^{\ast}$. As model average parameters become
closer to $\mathbf{w}^{\ast}$, the $L$-smoothness property implies
that $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}_{t}^{k})\|^{2}$
will become close to $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}^{\ast})\|^{2}$.
Therefore, there is no substantial difference between these assumptions
and assuming the bounds at $\mathbf{w}^{\ast}$ only. On the other
hand, some works such as \cite{haddadpour2019convergence,li2018federated}
assume bounds on the ``gradient diversity'' or ``dissimilarity'',
defined as $\frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$.
Such assumptions are more restrictive than our assumptions, because
they effectively require $\nabla F_{k}(\mathbf{w}^{\ast})\equiv0$ for all local
objectives. For such ``overparameterized'' settings, we will prove
that the convergence is in fact geometric.

>>>>

In this paper, we study the following federated learning problem:
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum\nolimits_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In $k$-th local device, there are $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. Each device only has access to its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumptions on the
data distributions of each local device. The local minimum
$\vw^{*}_k= \argmin_{\vw \in R^d} F_k(\vw)$ can be far from
the global minimum of \eq{\ref{eq:problem}}.

@@

\subsection{The Federated Averaging (FedAvg) Algorithm}
We first introduce the standard Federated Average (FedAvg) algorithm~\cite{mcmahan2016communication}.
FedAvg updates the model in each device by local Stochastic Gradient Descent (SGD) and send the latest model to the central server every $E$
steps. The central server conducts a weighted average over the model parameters
received from active devices and broadcasts the latest averaged model to all devices.
Formally, the updates of FedAvg at round $t$ is described as follows:
\begin{align}
\vv_{t+1}^{k} & =\vw_{t}^{k}-\alpha_{t}\vg_{t,k}\\
\mathbf{w}_{t+1}^{k} &=\left\{
\begin{array}{ll}
\mathbf{v}_{t+1}^{k} & \text { if } t+1 \notin \mathcal{I}_{E}, \\ 
\sum_{k \in \cS_{t+1}} \mathbf{v}_{t+1}^{k} & \text { if } t+1 \in \mathcal{I}_{E}
\end{array}\right.
\end{align}
where $\vw_t^k$ is the local model parameter maintained in the $k$-th device at the $t$-th iteration, $\vg_{t,k}:=\nabla F_{k}(\vw_{t}^{k},\xi_{t}^{k})$ is the stochastic gradient based on $\xi_{t}^{k}$, the data sampled from $k$-th device’s local data uniformly at random. $\mathcal{I}_{E}=\{E,2E,\dots\}$ is the set of global communication steps. We use $\mathcal{S}_{t+1}$ to represent the set of active devices at $t+1$.

Since federated learning usually involves an enormous amount of 
local devices, it is often more realistic to assume only a subset of 
local devices is active at each communication round (system heterogeneity). In this work,
we consider both the case of \textbf{full participation} where the model is
averaging over all devices at the communication round, i.e.,
$\mathbf{w}_{t+1}^{k} = \sum_{k=1}^N p_k \mathbf{v}_{t+1}^{k}$, and
the case of \textbf{partial participation} where $|\cS_{t+1}| < N$. 
With partial participation, $\cS_{t+1}$ is obtained by two types of
sampling schemes to simulate practical scenarios~\cite{li2019convergence}. For example, Sampling scheme I establishes $\cS_{t+1}$ by i.i.d. sampling the devices with probability $p_k$ with replacement. 
Both schemes
guarantee that gradient updates in FedAvg are unbiased stochastic versions of
updates in FedAvg with full participation. For 
more details on the notations and setup, please refer to Section~\ref{sec:app:notations} in the appendix.

@@

\subsection{Assumptions}
\label{sec:assumptions}
We make the following standard assumptions on the objective function $F_1,\dots, F_N$. Assumptions~\ref{ass:lsmooth} and~\ref{ass:stroncvx} are commonly satisfied by a range of popular objective functions, such as $\ell^{2}$ regularized logistic regression and cross-entropy loss functions.

@@

\textbf{Justfication for Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2}.} These standard assumptions have been made in many previous
works in federated learning\cite{yu2019parallel,li2019convergence,stich2018local}. We provide one more justification
for their generality: even though they require bounds for all $\mathbf{w}$,
they are not much stronger than requiring bounds at the global optimal
parameter $\mathbf{w}^{\ast}$. As model average parameters become
closer to $\mathbf{w}^{\ast}$, the $L$-smoothness property implies
that $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}_{t}^{k})\|^{2}$
will become close to $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}^{\ast})\|^{2}$.
Therefore, there is no substantial difference between these assumptions
and assuming the bounds at $\mathbf{w}^{\ast}$ only. On the other
hand, some works such as \cite{haddadpour2019convergence,li2018federated}
assume bounds on the ``gradient diversity'' or ``dissimilarity'',
defined as $\frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$.
Such assumptions are more restrictive than our assumptions, because
they effectively require $\nabla F_{k}(\mathbf{w}^{\ast})\equiv0$ for all local
objectives. For such ``overparameterized'' settings, we will prove
that the convergence is in fact geometric.
